\textbf{Evidence for a Change to Aerosol Fractions: In a Gobi Town}

### 2024-11-25 submission/Outline_20241121.Rmd
outline - title changed, abstract started, results drafted except the last figures

---
title: "**Spatio-temporal distinct patterns in variations of $PM_{10}$ and $PM_{2.5}$
  relative to the recent drivings of emission sources in Mongolia**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
  html_document:
    df_print: paged
csl: "mbio.csl"
fontsize: 11pt
geometry: margin=1.0in
editor_options:
  markdown:
    wrap: 72
---

@@

## Abstract (150 words)

Storyline:

1.  A new pattern is emerged
2.  Air quality in urban sites is episodically dictated by dust events
    in spring or late autumn, yet seasonally governed by anthropogenic
    emissions in winter. [Air quality is governed by natural dust
    emission, and anthropogenic emissions]
3.  With recent growing interest in urban life style, and combustion of
    coal/oyutolgoi for heating winter conditions results a highly
    increase in not only capital city but also towns
4.  In a result, spring coarse dust, plus winter fine pollutants
5.  spring coarse dust is immediately transported and deposited in the
    source area, whereas winter fine pollutants is permanently stayed in
    the source area due to stagnant atmosphere govern over entire
    country., perhaps floating in the near surface, deposits in the
    surface]
6.  Alarms, the Mongolian dust in the spring, optical properties might
    be shifted; this gives ... Gobi dust and sand storms has become
    tuiren, from the shoroon shuurga. which clearly requires the
    attention.
7.  r ratio shows ... emission source; dust might carry anthropogenic
    fine particulates as well.

@@

--> 3.1.2

@@

--> 3.2.1

### The emission patterns of interrelations among meteorological variables at the study sites
  - PMs varies with the wind speed and visibility
  - In general, three distinct patterns were resulted with PCA analysis, which is in consistent with temporal variation. explained with variables and changes in drivings (with PCA analysis)

[AND] To distinguish the emission driving variables, we demonstrated interrelations between(WS) and visibility (VIS) among particulate matters of PM10 and PM2.5 for study sites (figure_4).
[AND] Higher PM concentrations tend to correspond to lower wind speeds (smaller circles) and reduced visibility (darker blue points) in UB.
[AND] All Gobi sites show the largest variability in PM concentrations, with extreme outliers correlated with the WS primarily in the springs.

[AND] In comparison between PM10 increase with the PM2.5, as concentrations increase in PM10 concentration the ratio 1:1 is inscreased approximately 1.3 times for UB, and 2 times for all three Gobi sites. This increase in Gobi sites were further increased upto 5 times in ZU, and XXX times in SS sites showing the insufficient PM2.5 particulates in the those sites.

The Principal Component Analysis (PCA) biplot (Figure 7a) highlights the relationships among key variables, including PM10, PM2.5, wind speed (WS), wind direction (WD), visibility, relative humidity (r), and optical properties of aerosols (OPC). The first two principal components (Dim 1 and Dim 2) explain 66.52% of the total variance, with Dim 1 (44.36%) strongly associated with PM10 and PM2.5, indicating that these pollutants are major contributors to air quality variations. Visibility is inversely correlated with PM10 and PM2.5, suggesting that reduced visibility aligns with higher pollution levels. Wind speed and wind direction are positively aligned with Dim 2 (22.16%), reflecting their influence on pollutant dispersion.

The clustering of monitoring sites based on PM concentrations and geographic characteristics (Figure 7b) reveals distinct patterns. UB (urban background) is characterized by high PM concentrations (positive Dim 1 scores), while SS (suburban site) demonstrates low PM levels, clustering tightly along the negative Dim 1 axis. DZ (industrial site) exhibits substantial variability, with clusters extending into higher Dim 1 and Dim 2 ranges, reflecting the complex interplay of emission sources and meteorological factors. ZU (rural site) overlaps with SS but shows greater spread, indicating moderate pollution levels influenced by seasonal and localized factors. These results underscore the significant spatial and meteorological influences on air quality across the studied locations.

[BUT]

[THEREFORE] 3 clusters, supply is different?

###  The recent trends in concentrations of PMS and fine-coarse fractional changes at the sites
Figure 8 illustrates the trend analysis of PM10 and PM2.5 concentrations across four monitoring sites (UB, DZ, ZU, and SS) from 2009 to 2020. The time series demonstrates seasonal patterns and trends, with notable differences in particulate matter concentrations between the four seasons: Q1 (winter), Q2 (spring), Q3 (summer), and Q4 (autumn).

The p-values displayed on the figure indicate the statistical significance of trend changes in PM10 and PM2.5 levels for each season. At UB, significant negative trends are observed for Q1 and Q2 (p < 0.001). At DZ, a significant positive trend is observed for Q1 (R = 0.17, p < 0.001), indicating increasing particulate matter levels during this season.17, p < 0.001), while Q2 exhibits no significant trend (R = 0, p = 0.915)., highlighting a decreasing pattern in particulate concentrations during these periods over time. In contrast, Q3 and Q4 show weaker or nonsignificant trends (e.g., R = -0.03, p > 0.2), suggesting minimal changes in PM levels during the warmer months.

The industrial site (DZ) exhibits higher variability in PM levels, with periodic spikes indicative of localized emissions and industrial activities. Urban background (UB) consistently records elevated PM levels during Q1, driven by heating and combustion-related sources. Rural (ZU) and suburban (SS) sites generally display lower particulate concentrations. However, at ZU, a significant negative trend is observed in both PM10 and PM2.5 for Q1, suggesting a consistent reduction in particulate matter during winter months (may reduced transboundary traffic). At SS, a negative trend is observed in PM2.5 for Q2, reflecting a seasonal decline potentially linked to  reduced wind speed during spring.

The significant decreasing trends in PM concentrations in UB during colder months likely indicate improvements in emission control measures and a transition to more efficient heating practices. In contrast, a significant increasing trend is observed at DZ, particularly during Q1, suggesting rising emissions potentially associated with urbanized  household practices and other local activities in Mongolia's regional towns.

  - There are significant changes in time-series of PMs at 4 seasons
  - There any significant changes in ratio in the spring in respect to winter in DZ.
  - Close relationships was found between PM2.5 in winter and r values in the spring.

## Conclusions

In this study, we investigated the temporal variations of PM2.5 and PM10
concentrations at the 4 sites of rural and urban those located along the
the wind corridor. Three distinct variations has been detected.

Air quality in urban sites is episodically dictated by dust events in spring or late autumn, yet seasonally governed by anthropogenic emissions in winter.
Air quality in rural sites of SS and ZU is episodically dictated by dust events in spring or late autumn.
Air quality in rural sites of SS and ZU is episodically dictated by dust events in spring or late autumn.

A clear seasonal variations in the sites of UB and DZ is [Air quality is
governed by natural dust emission, and anthropogenic emissions]

>>>>

---
title: "**Evidence for a Change to Aerosol Fractions: In a Gobi Town**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
  html_document:
    df_print: paged
csl: "mbio.csl"
fontsize: 11pt
geometry: margin=1.0in
editor_options:
  markdown:
    wrap: 72
---

@@

## Abstract (150 words)
Driven by mining industrial progress, household coal combustion has increased greatly in Mongolia since the 2000s. Demographic evidence has revealed an ongoing reduction in rural-nomadic lifestyle and a rise of the population at urban areas, which may gradually extend a household major source of fine particulate matter (PM2.5) air pollution in winter as a heating. Together with observations that lifespan in various animal species is flexible and can be increased by genetic or pharmaceutical intervention, these results have led to suggestions that longevity may not be subject to strict, species-specific genetic constraints. Here, by analysing global demographic data, we show that improvements in survival with age tend to decline after age 100, and that the age at death of the world’s oldest person has not increased since the 1990s. Our results strongly suggest that the maximum lifespan of humans is fixed and subject to natural constraints.
Storyline:

1.  A new pattern is emerged
2.  Air quality in urban sites is episodically dictated by dust events
    in spring or late autumn, yet seasonally governed by anthropogenic
    emissions in winter. [Air quality is governed by natural dust
    emission, and anthropogenic emissions]
3.  With recent growing interest in urban life style, and combustion of
    coal/oyutolgoi for heating winter conditions results a highly
    increase in not only capital city but also towns
4.  In a result, spring coarse dust, plus winter fine pollutants
5.  spring coarse dust is immediately transported and deposited in the
    source area, whereas winter fine pollutants is permanently stayed in
    the source area due to stagnant atmosphere govern over entire
    country., perhaps floating in the near surface, deposits in the
    surface]
6.  Alarms, the Mongolian dust in the spring, optical properties might
    be shifted; this gives ... Gobi dust and sand storms has become
    tuiren, from the shoroon shuurga. which clearly requires the
    attention.
7.  r ratio shows ... emission source; dust might carry anthropogenic
    fine particulates as well.
    Spatio-temporal distinct patterns in variations of $PM_{10}$ and $PM_{2.5}$
      relative to the recent drivings of emission sources in Mongolia
\newpage

@@

@@

### The emission patterns of interrelations among meteorological variables at the study sites
add table, add r 1.3, 2 on figure_6, add supplement figures

  To identify the key factors influencing PM emissions, we examined the relationships between wind speed (WS), visibility (VIS), and particulate matter (PM10 and PM2.5) concentrations across the study sites (figure_6).
In UB during winter, elevated PM levels typically coincide with low wind speeds and reduced visibility (indicated by darker blue points). Notably, at DZ and ZU locations, high PM10 concentrations were observed during both low and high wind speed conditions in winter. All Gobi sites exhibit the greatest variation in PM concentrations during spring, with extreme outliers primarily associated with increased WS and reduced visibility below 10000 km. Similar findings were also revealed through Principal Component Analysis (PCA). The initial two principal components (Dim 1 and Dim 2) account for 66.52% of the total variance. Dim 1 shows a strong association with PM10 and PM2.5, while visibility demonstrates an inverse correlation, suggesting that reduced visibility corresponds to higher pollution levels. Wind speed and direction align positively with Dim 2 (22.16%), reflecting their impact on emissions.
When comparing PM10 aligns with PM2.5, the 1:1 ratio increases approximately 1.3 times for UB, 2 times for DZ, XXX for ZU, and XXX for SS as PM10 concentrations increase. This increase in Gobi sites further escalated up to 2-5 times in ZU, and XXX times in SS sites, indicating insufficient PM2.5 particulates in those locations, particularly at the ZU site. It is worth noting that instances where PM2.5 values exceed PM10 reflect equipment accuracy, correction errors, and higher sensitivity rates (ability to record emissions such as smoking near the sensor area). However, this discrepancy diminishes as PM concentrations increase, validating that higher observational records of PM2.5 compared to PM10 do not invalidate our results.  Furthermore, clustering of monitoring sites based on PM concentrations and geographic features (Figure 7b) reveals distinct patterns. UB (urban, capital city) is characterized by high PM concentrations (positive Dim 1 scores), while SS (rural, Gobi, town (site located in the prevailing wind above the town)) shows low PM levels, clustering tightly along the negative Dim 1 axis. DZ (rural, Gobi, town center) displays considerable variability, with clusters extending into higher Dim 1 and Dim 2 ranges, reflecting the complex interplay of emission sources and meteorological factors. ZU (rural, Gobi, village) overlaps with SS but exhibits greater spread, indicating moderate pollution levels influenced by seasonal and localized factors.
These findings highlight the complex interplay of spatial, meteorological, and local factors influencing air particulate matter concentrations across the studied locations and emphasizing the need for considering not only regional influences but also site-specific characteristics.

fig caption:  the Principal Component Analysis (PCA) bi-plot (Figure 7a) highlights the relationships among key variables, including PM10, PM2.5, wind speed (WS), wind direction (WD), visibility, ratio of PM2.5 to PM10 (r), and numbers of aerosols by optical particle counter (OPC). The first two principal components (Dim 1 and Dim 2) explain 66.52% of the total variance, with Dim 1 (44.36%) strongly associated with PM10 and PM2.5. Wind speed and wind direction are positively aligned with Dim 2 (22.16%), reflecting their influence on pollutant dispersion.

###  The recent trends in concentrations of PMS and fine-coarse fractional changes at the sites
add table, add trend figure: r, add relationship figure: r in spring and pm2.5 in winter

  Figure 8 illustrates the trend analysis of PM10 and PM2.5 concentrations across study sites from 2009 to 2020.
The time series demonstrates seasonal patterns and trends, with p-values of trend changes, seasonally: winter (Q1), spring (Q2), and summer-autumn (Q3). At UB, significant negative trends both in PM10 and PM2.5 concentrations are observed for winter and spring (Q1 and Q2, p < 0.001). At DZ, a significant positive trend in PM10 and PM2.5 concentrations is observed for winter (Q1, p < 0.001), indicating increasing particulate matter levels during this season.
At ZU, a significant negative trend is observed in both PM10 and PM2.5 for winter (Q1, p < 0.001), suggesting a consistent reduction in particulate matter during winter months (may reduced transboundary traffic associated with covid periods, resulted in declined PM10 concentrations; or data gaps). At SS, a negative trend is observed in PM2.5 for spring (Q2, p < 0.001), reflecting a seasonal decline potentially linked to  reduced wind speed during spring.
The significant decreasing trends in PM concentrations in UB during colder months likely indicate improvements in emission control measures and a transition to more efficient heating practices. In contrast, a significant increasing trend is observed at DZ, particularly during winter (Q1), suggesting rising emissions potentially associated with urbanized  household practices and other local activities in Mongolia's regional towns.

## Conclusions
In this study, we investigated the temporal variations of PM2.5 and PM10 concentrations at the 4 sites of rural and urban those located along the
the wind corridor.
Air particulate matter concentrations in urban-town sites of UB and DZ is episodically dictated by dust events in spring or late autumn, yet seasonally governed by anthropogenic emissions in winter. Air particulate matter concentrations in rural sites of SS and ZU is episodically dictated by dust events in spring or late autumn.

Three distinct variations has been detected.

1. It is evident of the new emission patterns in Mongolia.
2. Related to the winter emission patterns, fine particulates fraction in the spring is increased.
3. This alarms...
4. National level; meteorological impact is large. ... However, reduced... On the other hand, it is ... with the towns. This points that air quality will be poor whether it is changed fuel, .. unless to change heating system. Therefore, it is not the reason to move the capital city. Only solution is to change the heating system, do not burn any type of the coal.

### 2024-11-25 submission/Outline_20241121.tex
outline - title changed, abstract started, results drafted except the last figures

\usepackage{pdfpages}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Spatio-temporal distinct patterns in variations of PM\_\{10\} and PM\_\{2.5\} relative to the recent drivings of emission sources in Mongolia},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Spatio-temporal distinct patterns in variations of
\(PM_{10}\) and \(PM_{2.5}\) relative to the recent drivings of emission
sources in Mongolia}}
\author{}
\date{\vspace{-2.5em}}

@@

Storyline:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A new pattern is emerged
\item
  Air quality in urban sites is episodically dictated by dust events in
  spring or late autumn, yet seasonally governed by anthropogenic
  emissions in winter. {[}Air quality is governed by natural dust
  emission, and anthropogenic emissions{]}
\item
  With recent growing interest in urban life style, and combustion of
  coal/oyutolgoi for heating winter conditions results a highly increase
  in not only capital city but also towns
\item
  In a result, spring coarse dust, plus winter fine pollutants
\item
  spring coarse dust is immediately transported and deposited in the
  source area, whereas winter fine pollutants is permanently stayed in
  the source area due to stagnant atmosphere govern over entire
  country., perhaps floating in the near surface, deposits in the
  surface{]}
\item
  Alarms, the Mongolian dust in the spring, optical properties might be
  shifted; this gives \ldots{} Gobi dust and sand storms has become
  tuiren, from the shoroon shuurga. which clearly requires the
  attention.
\item
  r ratio shows \ldots{} emission source; dust might carry anthropogenic
  fine particulates as well.
\end{enumerate}

@@

--\textgreater{} 3.1.2

@@

--\textgreater{} 3.2.1

@@

\begin{itemize}
\tightlist
\item
  PMs varies with the wind speed and visibility
\item
  In general, three distinct patterns were resulted with PCA analysis,
  which is in consistent with temporal variation. explained with
  variables and changes in drivings (with PCA analysis)
\end{itemize}

{[}AND{]} To distinguish the emission driving variables, we demonstrated
interrelations between(WS) and visibility (VIS) among particulate
matters of PM10 and PM2.5 for study sites (figure\_4). {[}AND{]} Higher
PM concentrations tend to correspond to lower wind speeds (smaller
circles) and reduced visibility (darker blue points) in UB. {[}AND{]}
All Gobi sites show the largest variability in PM concentrations, with
extreme outliers correlated with the WS primarily in the springs.

{[}AND{]} In comparison between PM10 increase with the PM2.5, as
concentrations increase in PM10 concentration the ratio 1:1 is
inscreased approximately 1.3 times for UB, and 2 times for all three
Gobi sites. This increase in Gobi sites were further increased upto 5
times in ZU, and XXX times in SS sites showing the insufficient PM2.5
particulates in the those sites.

The Principal Component Analysis (PCA) biplot (Figure 7a) highlights the
relationships among key variables, including PM10, PM2.5, wind speed
(WS), wind direction (WD), visibility, relative humidity (r), and
optical properties of aerosols (OPC). The first two principal components
(Dim 1 and Dim 2) explain 66.52\% of the total variance, with Dim 1
(44.36\%) strongly associated with PM10 and PM2.5, indicating that these
pollutants are major contributors to air quality variations. Visibility
is inversely correlated with PM10 and PM2.5, suggesting that reduced
visibility aligns with higher pollution levels. Wind speed and wind
direction are positively aligned with Dim 2 (22.16\%), reflecting their
influence on pollutant dispersion.

The clustering of monitoring sites based on PM concentrations and
geographic characteristics (Figure 7b) reveals distinct patterns. UB
(urban background) is characterized by high PM concentrations (positive
Dim 1 scores), while SS (suburban site) demonstrates low PM levels,
clustering tightly along the negative Dim 1 axis. DZ (industrial site)
exhibits substantial variability, with clusters extending into higher
Dim 1 and Dim 2 ranges, reflecting the complex interplay of emission
sources and meteorological factors. ZU (rural site) overlaps with SS but
shows greater spread, indicating moderate pollution levels influenced by
seasonal and localized factors. These results underscore the significant
spatial and meteorological influences on air quality across the studied
locations.

{[}BUT{]}

{[}THEREFORE{]} 3 clusters, supply is different?

@@

Figure 8 illustrates the trend analysis of PM10 and PM2.5 concentrations
across four monitoring sites (UB, DZ, ZU, and SS) from 2009 to 2020. The
time series demonstrates seasonal patterns and trends, with notable
differences in particulate matter concentrations between the four
seasons: Q1 (winter), Q2 (spring), Q3 (summer), and Q4 (autumn).

The p-values displayed on the figure indicate the statistical
significance of trend changes in PM10 and PM2.5 levels for each season.
At UB, significant negative trends are observed for Q1 and Q2 (p
\textless{} 0.001). At DZ, a significant positive trend is observed for
Q1 (R = 0.17, p \textless{} 0.001), indicating increasing particulate
matter levels during this season.17, p \textless{} 0.001), while Q2
exhibits no significant trend (R = 0, p = 0.915)., highlighting a
decreasing pattern in particulate concentrations during these periods
over time. In contrast, Q3 and Q4 show weaker or nonsignificant trends
(e.g., R = -0.03, p \textgreater{} 0.2), suggesting minimal changes in
PM levels during the warmer months.

The industrial site (DZ) exhibits higher variability in PM levels, with
periodic spikes indicative of localized emissions and industrial
activities. Urban background (UB) consistently records elevated PM
levels during Q1, driven by heating and combustion-related sources.
Rural (ZU) and suburban (SS) sites generally display lower particulate
concentrations. However, at ZU, a significant negative trend is observed
in both PM10 and PM2.5 for Q1, suggesting a consistent reduction in
particulate matter during winter months (may reduced transboundary
traffic). At SS, a negative trend is observed in PM2.5 for Q2,
reflecting a seasonal decline potentially linked to reduced wind speed
during spring.

The significant decreasing trends in PM concentrations in UB during
colder months likely indicate improvements in emission control measures
and a transition to more efficient heating practices. In contrast, a
significant increasing trend is observed at DZ, particularly during Q1,
suggesting rising emissions potentially associated with urbanized
household practices and other local activities in Mongolia's regional
towns.

\begin{itemize}
\tightlist
\item
  There are significant changes in time-series of PMs at 4 seasons
\item
  There any significant changes in ratio in the spring in respect to
  winter in DZ.
\item
  Close relationships was found between PM2.5 in winter and r values in
  the spring.
\end{itemize}

@@

In this study, we investigated the temporal variations of PM2.5 and PM10
concentrations at the 4 sites of rural and urban those located along the
the wind corridor. Three distinct variations has been detected.

Air quality in urban sites is episodically dictated by dust events in
spring or late autumn, yet seasonally governed by anthropogenic
emissions in winter. Air quality in rural sites of SS and ZU is
episodically dictated by dust events in spring or late autumn. Air
quality in rural sites of SS and ZU is episodically dictated by dust
events in spring or late autumn.

A clear seasonal variations in the sites of UB and DZ is {[}Air quality
is governed by natural dust emission, and anthropogenic emissions{]}

>>>>

\usepackage{pdfpages}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evidence for a Change to Aerosol Fractions: In a Gobi Town},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Evidence for a Change to Aerosol Fractions: In a Gobi
Town}}
\author{}
\date{\vspace{-2.5em}}

@@

Driven by mining industrial progress, household coal combustion has
increased greatly in Mongolia since the 2000s. Demographic evidence has
revealed an ongoing reduction in rural-nomadic lifestyle and a rise of
the population at urban areas, which may gradually extend a household
major source of fine particulate matter (PM2.5) air pollution in winter
as a heating. Together with observations that lifespan in various animal
species is flexible and can be increased by genetic or pharmaceutical
intervention, these results have led to suggestions that longevity may
not be subject to strict, species-specific genetic constraints. Here, by
analysing global demographic data, we show that improvements in survival
with age tend to decline after age 100, and that the age at death of the
world's oldest person has not increased since the 1990s. Our results
strongly suggest that the maximum lifespan of humans is fixed and
subject to natural constraints. Storyline:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A new pattern is emerged
\item
  Air quality in urban sites is episodically dictated by dust events in
  spring or late autumn, yet seasonally governed by anthropogenic
  emissions in winter. {[}Air quality is governed by natural dust
  emission, and anthropogenic emissions{]}
\item
  With recent growing interest in urban life style, and combustion of
  coal/oyutolgoi for heating winter conditions results a highly increase
  in not only capital city but also towns
\item
  In a result, spring coarse dust, plus winter fine pollutants
\item
  spring coarse dust is immediately transported and deposited in the
  source area, whereas winter fine pollutants is permanently stayed in
  the source area due to stagnant atmosphere govern over entire
  country., perhaps floating in the near surface, deposits in the
  surface{]}
\item
  Alarms, the Mongolian dust in the spring, optical properties might be
  shifted; this gives \ldots{} Gobi dust and sand storms has become
  tuiren, from the shoroon shuurga. which clearly requires the
  attention.
\item
  r ratio shows \ldots{} emission source; dust might carry anthropogenic
  fine particulates as well. Spatio-temporal distinct patterns in
  variations of \(PM_{10}\) and \(PM_{2.5}\) relative to the recent
  drivings of emission sources in Mongolia \newpage
\end{enumerate}

@@

@@

@@

add table, add r 1.3, 2 on figure\_6, add supplement figures

To identify the key factors influencing PM emissions, we examined the
relationships between wind speed (WS), visibility (VIS), and particulate
matter (PM10 and PM2.5) concentrations across the study sites
(figure\_6). In UB during winter, elevated PM levels typically coincide
with low wind speeds and reduced visibility (indicated by darker blue
points). Notably, at DZ and ZU locations, high PM10 concentrations were
observed during both low and high wind speed conditions in winter. All
Gobi sites exhibit the greatest variation in PM concentrations during
spring, with extreme outliers primarily associated with increased WS and
reduced visibility below 10000 km. Similar findings were also revealed
through Principal Component Analysis (PCA). The initial two principal
components (Dim 1 and Dim 2) account for 66.52\% of the total variance.
Dim 1 shows a strong association with PM10 and PM2.5, while visibility
demonstrates an inverse correlation, suggesting that reduced visibility
corresponds to higher pollution levels. Wind speed and direction align
positively with Dim 2 (22.16\%), reflecting their impact on emissions.
When comparing PM10 aligns with PM2.5, the 1:1 ratio increases
approximately 1.3 times for UB, 2 times for DZ, XXX for ZU, and XXX for
SS as PM10 concentrations increase. This increase in Gobi sites further
escalated up to 2-5 times in ZU, and XXX times in SS sites, indicating
insufficient PM2.5 particulates in those locations, particularly at the
ZU site. It is worth noting that instances where PM2.5 values exceed
PM10 reflect equipment accuracy, correction errors, and higher
sensitivity rates (ability to record emissions such as smoking near the
sensor area). However, this discrepancy diminishes as PM concentrations
increase, validating that higher observational records of PM2.5 compared
to PM10 do not invalidate our results. Furthermore, clustering of
monitoring sites based on PM concentrations and geographic features
(Figure 7b) reveals distinct patterns. UB (urban, capital city) is
characterized by high PM concentrations (positive Dim 1 scores), while
SS (rural, Gobi, town (site located in the prevailing wind above the
town)) shows low PM levels, clustering tightly along the negative Dim 1
axis. DZ (rural, Gobi, town center) displays considerable variability,
with clusters extending into higher Dim 1 and Dim 2 ranges, reflecting
the complex interplay of emission sources and meteorological factors. ZU
(rural, Gobi, village) overlaps with SS but exhibits greater spread,
indicating moderate pollution levels influenced by seasonal and
localized factors. These findings highlight the complex interplay of
spatial, meteorological, and local factors influencing air particulate
matter concentrations across the studied locations and emphasizing the
need for considering not only regional influences but also site-specific
characteristics.

fig caption: the Principal Component Analysis (PCA) bi-plot (Figure 7a)
highlights the relationships among key variables, including PM10, PM2.5,
wind speed (WS), wind direction (WD), visibility, ratio of PM2.5 to PM10
(r), and numbers of aerosols by optical particle counter (OPC). The
first two principal components (Dim 1 and Dim 2) explain 66.52\% of the
total variance, with Dim 1 (44.36\%) strongly associated with PM10 and
PM2.5. Wind speed and wind direction are positively aligned with Dim 2
(22.16\%), reflecting their influence on pollutant dispersion.

@@

add table, add trend figure: r, add relationship figure: r in spring and
pm2.5 in winter

Figure 8 illustrates the trend analysis of PM10 and PM2.5 concentrations
across study sites from 2009 to 2020. The time series demonstrates
seasonal patterns and trends, with p-values of trend changes,
seasonally: winter (Q1), spring (Q2), and summer-autumn (Q3). At UB,
significant negative trends both in PM10 and PM2.5 concentrations are
observed for winter and spring (Q1 and Q2, p \textless{} 0.001). At DZ,
a significant positive trend in PM10 and PM2.5 concentrations is
observed for winter (Q1, p \textless{} 0.001), indicating increasing
particulate matter levels during this season. At ZU, a significant
negative trend is observed in both PM10 and PM2.5 for winter (Q1, p
\textless{} 0.001), suggesting a consistent reduction in particulate
matter during winter months (may reduced transboundary traffic
associated with covid periods, resulted in declined PM10 concentrations;
or data gaps). At SS, a negative trend is observed in PM2.5 for spring
(Q2, p \textless{} 0.001), reflecting a seasonal decline potentially
linked to reduced wind speed during spring. The significant decreasing
trends in PM concentrations in UB during colder months likely indicate
improvements in emission control measures and a transition to more
efficient heating practices. In contrast, a significant increasing trend
is observed at DZ, particularly during winter (Q1), suggesting rising
emissions potentially associated with urbanized household practices and
other local activities in Mongolia's regional towns.

@@

In this study, we investigated the temporal variations of PM2.5 and PM10
concentrations at the 4 sites of rural and urban those located along the
the wind corridor. Air particulate matter concentrations in urban-town
sites of UB and DZ is episodically dictated by dust events in spring or
late autumn, yet seasonally governed by anthropogenic emissions in
winter. Air particulate matter concentrations in rural sites of SS and
ZU is episodically dictated by dust events in spring or late autumn.

Three distinct variations has been detected.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  It is evident of the new emission patterns in Mongolia.
\item
  Related to the winter emission patterns, fine particulates fraction in
  the spring is increased.
\item
  This alarms\ldots{}
\item
  National level; meteorological impact is large. \ldots{} However,
  reduced\ldots{} On the other hand, it is \ldots{} with the towns. This
  points that air quality will be poor whether it is changed fuel, ..
  unless to change heating system. Therefore, it is not the reason to
  move the capital city. Only solution is to change the heating system,
  do not burn any type of the coal.
\end{enumerate}
